### 2017-02-25 2017mergingmerges/section/XX-Processes.tex
squeeze fused figure into columns

%!TEX root = ../Main.tex
\section{Processes, Combinators and Operators}
\label{s:Processes}

A \emph{process} in our system is a simple imperative program with a local heap. A process pulls data from an arbitrary number of input streams and pushes result values to at least one output stream. The process language is an intermediate representation we use when fusing the overall dataflow network. A \emph{combinator} is a template for a particular \emph{process} which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process. Each process implements a logical \emph{operator}, so we use ``operator'' when describing the values being computed, but ``process'' when referring to the implementation.

@@

The initial instruction @(pull sIn1 v A1 {})@ pulls the next element from the stream @sIn1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@. The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to update anything yet we leave it empty.

Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to the instruction at label @A3@. We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.

@@

The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}. The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@. The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream. The process starts by pulling from each input streams. It then compares the two pulled values, and pushes the smaller of the values to the output stream. The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.

As this process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output. For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @sIn2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.

% -----------------------------------------------------------------------------
\subsection{Fusion}
Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both. For example, suppose we need a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}. The result will be a machine that computes the result of both the @group@ and @merge@ if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@. For ease of comparison we will assume the parameters in each combinator are instantiated by arguments with the same names.

@@

We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream. For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two input states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @sIn1@ and @sIn2@ map to `none'.

From the joint state @F0@, both of the input machines then need to pull from stream @sIn1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@. In the result machine this is managed by storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@. For this we use the @jump@ instruction which transitions between states without affecting any of the input or output streams.

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine. In Fig.~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.

@@

Once the result machine has arrived in the joint state @F3@, this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively.
Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines.
From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to its output stream @sOut1@, or to just move on to the next value from its input. From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@. In the result, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
At @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.

Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
Because the stream @sIn22@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.

Note that we could construct the fused machine in multiple ways. One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch. By construction, the predicate used in the branch only refers to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine. As the set of variables does not overlap, either ordering is correct. For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.

@@

Fig.~\ref{fig:Process:Fused} shows the entire result of fusing @group@ and @merge@ together. There are other rules for handling different combinations of instructions, but we defer the details to \S\ref{s:Fusion}. The result process has two inputs, @sIn1@ and @sIn2@, which correspond to the original input streams. Of the two outputs, @sOut1@ is the @group@ processes output stream, while @sOut2@ is the @merge@ processes output stream.

Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequental result process that implements both. We have chosen a \emph{specific sequential order} in which to interleave instructions that implement both original processes. As with the Flow Fusion system of \cite{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time. However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss further in \S\ref{s:FutureWork}.

@@

Finally, although the result process has a single shared heap, the bindings are guaranteed not to interfere. When we instantiate combinators to create the original processes we use fresh names at that point. The stream buffer variables we aditionally introduce for coordination are freshly created during fusion.

@@

\newcommand\annot[5]{
  \tiny ((#1,   \> \tiny \{sIn1 =      #2\}), 
                \> \tiny (#3, \> \tiny \{sIn1 =      #4, \> \tiny sIn2 =      #5\}))
}
\newcommand\icase[7]{
 \tt{#1} \> \tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \tt{#5} \{\tt{#6}\} \> \tiny #1 \> \tiny = #7 \\
}
\newcommand\instr[5]{
 \tt{#1}\>\tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \> \tiny #1 \> \tiny = #5 \\
}
\begin{tabbing}
@  @ \=
@, F17@  \= = @case (f || (l /= v)) @
         \= @F17{x1 = b1, f = F}@ \= @F17{ }@
@      @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{s1 = pending\}), \= \tiny (B0, \= \tiny \{s1 = pending, \= \tiny s2 = pending\})) \kill
\> @{@ \instr{F0}{pull sIn1 b1}{F1}{}
      {\annot{A0}{none}{B0}{none}{none}}

>>>>

%!TEX root = ../Main.tex
\section{Processes, Machines, Combinators and Operators}
\label{s:Processes}

A \emph{process} in our system is a simple imperative program with a local heap. A process pulls data from an arbitrary number of input streams and pushes result values to at least one output stream. The process language is an intermediate representation we use when fusing the overall dataflow network. When describing the fusion transform we describe the control flow of the process as a state machine, hence Machine Fusion.

A \emph{combinator} is a template for a particular process which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process. Each process implements a logical \emph{operator} --- so we use ``operator'' when describing the values being computed, but ``process'' and ``machine'' when referring to the implementation.

@@

The initial instruction @(pull sIn1 v A1 {})@ pulls the next element from the stream @sIn1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@. The empty set @{}@ after the target label @A1@ can be used to update heap variables, but as we do not need to update anything yet we leave it empty.

Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to @A3@. We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) for whether this is the first element.

@@

The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}. The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@. The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream. The process starts by pulling from each of the input streams. It then compares the two pulled values, and pushes the smaller of the values to the output stream. The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.

As this the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output. For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output. After pushing the last value $4$, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.

% -----------------------------------------------------------------------------
\subsection{Fusion}
Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both. For example, suppose we need a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}. The result will be a machine that computes the result of both the @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@. For ease of comparison we will assume the parameters in each combinator are instantiated by arguments with the same names.

@@

We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream. For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two input states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

From the joint state @F0@, both of the input machines then need to pull from stream @sIn1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@. In the result machine this is managed by first storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@. For this we attach updates to a @jump@ instruction, which otherwise transitions between states without affecting any of the input or output streams.

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transition from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine. In Fig.~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.

@@

Once the result machine has arrived in the joint state @F3@, this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively. The left of Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines. From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to its output stream @sOut1@, or to just move on to the next value from its input. From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@. In the result machine, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively. At @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.

Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
Because the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.

Note that we could construct the fused result machine in several ways. One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch. By construction, the predicate used in the branch refers only to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine. As the set of variables does not overlap, either ordering is correct. For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.

@@

Fig.~\ref{fig:Process:Fused} shows the end result of fusing @group@ and @merge@ together. There are other rules for handling different combinations of instructions, but we defer the details to \S\ref{s:Fusion}. The result process has two inputs, @sIn1@ and @sIn2@, which correspond to the original input streams. It also has two outputs, @sOut1@ which is the @group@ processes output stream, while @sOut2@ is the @merge@ processes output stream.

Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequental result process that implements both. We have chosen a \emph{specific sequential order} in which to interleave instructions that implement both original processes. As with the Flow Fusion system of \citet{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time. However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss further in \S\ref{s:FutureWork}.

@@

Finally, although the result process has a single shared heap, the bindings are guaranteed not to interfere. When we instantiated combinators to create the original processes we introduced fresh names at that point. The stream buffer variables we aditionally introduced for coordination were freshly created during fusion.

@@

\newcommand\annot[5]{
  \tiny ((#1,   \> \tiny \{sIn1 =      #2\}), 
                \> \tiny (#3, \> \tiny \{sIn1 =      #4, \> \tiny sIn2 =      #5\}))
}
\newcommand\icase[7]{
 \tt{#1} \> \tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \tt{#5} \{\tt{#6}\} \> \tiny #1 \> \tiny = #7 \\
}
\newcommand\instr[5]{
 \tt{#1}\>\tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \> \tiny #1 \> \tiny = #5 \\
}
\begin{tabbing}
@  @ \=
@, F17@  \= = @case (f || (l /= v)) @
         \= @F17{}         @ \= @F17{ }@
@   @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
\> @{@ \instr{F0}{pull sIn1 b1}{F1}{}
      {\annot{A0}{none}{B0}{none}{none}}
